A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-06-09 doc/large_strain.tex
change how we mention Clevenger paper
cite Martin' work at the beginning of MF section

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper currently in preparation that deals with small strain linear elasticity \cite{Clevenger2018}.
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, that are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high}.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

>>>>

To the best of our knowledge, matrix-free methods are not widely adopted within the solid mechanics community.
We are only aware of a single paper that deals with small strain linear elasticity \cite{Clevenger2018}.
One possible reason is that the tangent operators\footnote{Here and
below 'operator' denotes a finite-dimensional linear operator defined on the vector space $\mathbb R^N$, where $N$ is the number of unknown degrees of freedom in the FE discretization.
Although any linear mapping on a finite-dimensional space is representable by a matrix, we adopt the `operator' term to avoid confusion between its matrix-free and matrix-based numerical implementation.
} (stemming from linearization of the nonlinear balance equations) are
more elaborate as compared to the operators used often in both linear and non-linear fluid mechanics, and it is not obvious whether matrix-free methods can be advantageous in this case.
Another reason could be the frequent use of lower-order elements in solid mechanics due to lacking smoothness of the underlying solution.
The finite element method (FEM) with linear or quadratic elements (that reduce the potential for locking), or mixed or enhanced formulations (that avoid locking) are often adopted in the solid mechanics community,
whereas higher order elements are rarely employed.
This can potentially reduce the advantage of matrix-free methods using sum-factorization techniques, that are generally more competitive for higher order elements \cite{kronbichler12,kronbichler2017fast,muthing2017high}.
However, we note that, even with linear FEM, large scale computations with $10^{12}$ unknowns inevitably require a matrix-free approach as there is simply not enough memory to store the sparse tangent matrix \cite{Gmeiner2016}. Therefore, we believe that for large scale computations in solid mechanics it is crucial to consider matrix-free approaches and develop efficient multigrid solvers.

@@

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case, the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches \cite{kronbichler12, kronbichler2017fast} where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.
